---
title: "Chapter 4: Geocentric models"
format: html
fig-cap-location: margin
reference-location: margin
citation-location: margin
editor_options: 
  chunk_output_type: console
---

```{r}
#| output: false
library(rethinking)
```

This chapter introduces linear regression as a bayesian procedure.

## Why normal distribution are normal?

### Normal by addition 

Let's simulate an experiment. 
A coin is flipped and the person takes a random step 
between 0 and 1 meter in the indicated direction:
left if the coin is up, right otherwise.
We repeat the process for 16 steps and 1,000 peoples.

```{r}
#| fig-height: 3
par(mfrow=c(1,3))
for (n_steps in c(4, 16, 50)){
    position <- replicate(1000, sum(runif(n_steps, -1, 1)))
    hist(position, main=paste(n_steps, " steps"))
}
```

Any process that adds together random values from the same distribution 
converges to a normal.
Often the convergence is often fast.

### Normal by multiplication

Here's another way to get a normal distribution.
Assume that the growth of an organism is influenced by 12 loci,
each with several alleles that code for more growth.
As all loci interact together, the growth is increased by a percentage.

```{r}
growth <- replicate(10000, prod(1+runif(12,0,0.1)))
dens(growth, norm.comp=TRUE, xlab="Growth")
```

What's happening is that multiplying small numbers (close to 1), 
is approximately the same as addition.

$$ 1.1 \times 1.1 = (1+0.1)(1+0.1) = 1 + 0.2 + 0.01 \simeq 1.2 $$
The smaller the effect of each locus,
the better the additive approximation is.

```{r}
#| fig-height: 4
par(mfrow=c(1,2))
big <- replicate(10000, prod(1+runif(12,0,0.5)))
small <- replicate(10000, prod(1+runif(12,0,0.01)))
dens(big, norm.comp=TRUE, xlab="Growth", main="Big effect")
dens(small, norm.comp=TRUE, xlab="Growth", main="Small effect")
```

### Normal by log-multiplication

Large deviates multiplied do not produce Gaussian distributions, 
but they tend to produce Gaussian distributions on log scale.

```{r}
big_log <- replicate(10000, log(prod(1+runif(12,0,0.5))))
dens(big_log, norm.comp=TRUE, xlab="Growth", main="Big effect (log)")
```

We get a normal distribution again because $\log(ab) = \log(a) + \log(b)$.

### Using Gaussian distribution 

Gaussian distributions are a member of the *exponential family*. 
All the members of this family are important for working ins science, 
because they populate our world.
The Gaussian distribution is the shape that can be realized in the largest 
number of ways and does not introduce any new assumptions, 
other than the measure as a finite variance.

:::{.callout-note}
## Gaussian probability density

$$ p(y|\mu,\sigma) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp(- \frac{(y-\mu)^2}{2\sigma^2}) $$
Sometimes instead of $\sigma$ the precision $\tau = \sigma^{-2}$ is used.
:::